Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-01-26 papers/LOPe/coarray-extensions.tex
Almost finished with CAFe part.

So far --- except for the \texttt{HALO} attribute --- this is totally standard Fortran (2008).  For CAFe,
we extend the concept of Fortran images to be hierachical.  By this we mean that each image \emph{may}
have a subimage (or subimages), but this subimage is not visible to other normal Fortran images.
Subimages also execute differently than normal images.  Subimages are task based while images all
execute a single program but with different (multiple) data (SPMD).  A task can be given to a subimage,
but execution on the subimage terminates once the task is finished.  Memory on a subimage, however, is
permanent and must be specifically allocated and deallocated.

One obtains a subimage by executing,
\begin{verbatim}
   device = get_subimage(1)          ! THIS_IMAGE() if device unavailable
\end{verbatim}

\begin{verbatim}

   !! Allocate memory on subimage if available
   !  ----------------------------------------
   if (device /= THIS_IMAGE()) then
      allocate(U[device], HALO_SRC=U)   [[device]]
   end if

   !! Initialize and transfer memory
   !  ------------------------------
   do j1h = 0, J
      U1h(j1h) = j1h
   end do
   U1h[device] = U1h

   do i = 1, nLoops
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace(U(i,j)[device])
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do

   deallocate(U)
   if (device /= THIS_IMAGE()) then
      deallocate(U[device])   [[device]]
   end if

end program
\end{verbatim}

>>>>

So far --- except for the \texttt{HALO} attribute --- this is totally standard Fortran (2008).  For CAFe,
we extend the concept of Fortran images to be hierachical.  By this we mean that each image \emph{may}
have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
Subimages also execute differently than normal images.  Subimages are task based while images all
execute a single program but with different (multiple) data (SPMD).  A task can be given to a subimage,
but execution on the subimage terminates once the task is finished.  Memory on a subimage is
permanent, however, and must be specifically allocated and deallocated.

One obtains a subimage by executing a function call,
\begin{verbatim}
   device = get_subimage(1)
\end{verbatim}
where the integer argument representing one of attached hardware devices.  If the function
fails (e.g., the requested device is unavailable) it returns the image number,
\texttt{this\_image()}, of the process that is executing the current program.  Returning
the current image allows program execution to proceed correctly even if there are no
attached devices.

Once a subimage is obtained, memory on the device may be allocated,
\begin{verbatim}
  if (device /= this_image()) then
    allocate(U[device], HALO_SRC=U)   [[device]]
  end if
\end{verbatim}
There are four points to note regarding this memory allocation: 1. Memory is only
allocated if a subimage has been obtained; 2. The location where memory is allocated is
denoted by the regular coarray notation \texttt{U[device]}; 3. The allocated size and halo
attribute of the new array are assumed from the previously allocated local array
\texttt{U} via the notation \texttt{HALO\_SRC=U} (using \texttt{HALO\_SRC} will also copy
the memory to the device); and finally 4. The allocation itself is \emph{executed} on the
subimage device with the notation \texttt{[[device]]}.

Fortran uses square bracket notation \texttt{[image]} to specify on what process the
memory reference is physically located.  Square brackets are a visual clue to the
programmer that the memory reference may be remote and therefore potentially suffer a
performance penalty.  CAFe extends this by employing double-bracket notation to indicate
possibly \emph{remote subimage execution}.

Execution of the Laplace task (defined previously) is done using the Fortran \texttt{do
concurrent} construct as shown below,
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace(U(i,j)[device])
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The \texttt{local element} of the array (as
described above in reference to the definition of the \texttt{CONCURRENT} procedure
\texttt{Laplace} is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.

The transfer of halo memory is necessary after each completion of the do concurrent loop.
Completion of a do concurrent construct indicates that all executing threads (if a threading
model is used by the compiler) have completed and that all memory on the executing subimage is
in a consistent state.

Finally, deallocation of memory is similar to memory allocation:
\begin{verbatim}
   deallocate(U)
   if (device /= this_image()) then
      deallocate(U[device])   [[device]]
   end if
\end{verbatim}
